---
title: "Week 2 Assignment"
author: "Ricardo de la Torre"
date: "1/6/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is the Markdown document for the Week 2 Assignment on the Reproducible Research for Data Science.

## Step 1: What is mean total number of steps taken per day?

First, we load the data into a variable, omitting the NA information.

```{r load data, echo=TRUE}

activity <- read.csv("activity.csv")

tidy_activity <- na.omit(activity)

```

We need to generate an histogram with the total number of steps taken each day. We first generate a new entity with the sum, mean and median of steps per day, using dplyr package.

```{r step1 generate, echo=TRUE, results="hide", message=FALSE}

library(dplyr)

new_activity <- tidy_activity %>% 
   group_by(date) %>% 
   summarise(SumSteps = sum(steps))

```

Using the ggplot2 system, we generate the histogram for the total number of steps per day.

```{r step1 plot, echo=TRUE, message=FALSE, warning=FALSE}

library(ggplot2)

ggplot(new_activity, aes(SumSteps)) + geom_histogram(binwidth = 5000)

```

Finally, we calculate the mean and median of the total of steps

```{r step1 show, echo=TRUE}

#disable scientific notation
options(scipen=999)

meanStep <- mean(new_activity$SumSteps)
medianStep <- median(new_activity$SumSteps)


```

* The __mean__ total number of steps taken per day: _`r meanStep`_
* The __median__ total number of steps taken per day: _`r medianStep`_

## Step 2: What is the average daily activity pattern?

First, we have to get the information asked by twisting the activity data in a proper way. As requested, we have to plot the data with the x-axis as 5-minute interval and the average number of steps taken across all days in the y-axis.

```{r step2 generate, echo=TRUE, results="hide", message=FALSE}

avg_activity <- tidy_activity %>% 
   group_by(interval) %>% 
   summarise(AvgSteps = mean(steps))

```

The time series plot resulting with this information is the stated below, plotted using ggplot

```{r step2 plot, echo=TRUE, message=FALSE, warning=FALSE}

ggplot(avg_activity, aes(interval, AvgSteps)) + geom_line()

```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r step2 show, echo=TRUE}

maxIdx = which.max(avg_activity$AvgSteps)
maxInterval = avg_activity[maxIdx, 'interval']
maxAvgStep = avg_activity[maxIdx, 'AvgSteps']


```

* On average across all the days in the dataset, __`r maxInterval`__ 5-minute interval contains maximum number of steps (_`r maxAvgStep`_ steps)

## Step 3: Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r step3 part1, echo=TRUE}

totalNA = sum(is.na(activity$steps))

```
* The total number of missing values in the dataset: __`r totalNA`__

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

I opted for the strategy of filling all of missing values with the mean of 5-minute interval

``` {r step3 part2, warning=FALSE}

#get the NA index 
naIdx = which(is.na(activity$steps))
naInterval = activity[naIdx, 3] 

#get the corrosponding intervale
fillSteps = sapply(naInterval, function(x) { avg_activity[(avg_activity$interval==x), 2]})

#create a new dataset that is equal to the orginal dataset
# BUT with the missing data filled in 
dataNew = activity
dataNew[naIdx, 'steps'] = fillSteps 

#fill in missing data
#again, calculate the average of steps w.r.t. the time interval
sumStepNew = aggregate(steps~date, dataNew, sum)

#plot graph

ggplot(sumStepNew, aes(steps)) + geom_histogram(binwidth = 5000)

```

We then calculate the mean and median values with the addition of the new missing values

``` {r step3 part3}

#calculate the mean and median total number of steps taken per day
meanStepNew = mean(sumStepNew$steps)
medianStepNew = median(sumStepNew$steps)

```

* The __mean__ total number of steps taken per day: _`r meanStepNew`_
* The __median__ total number of steps taken per day: _`r medianStepNew`_

The mean number of steps taken per day are the same (`r meanStepNew`). 
But, the median number of steps taken per day are slightly different (before filling missing data: `r medianStep`, after filling missing data: `r medianStepNew`). 
It is probably due to filling missing values by the mean of 5 minute interval.

## Step 4: Are there differences in activity patterns between weekdays and weekends?

First, we need to create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

``` {r step4 part1}

dataNew$date = as.Date(as.character(dataNew$date), "%Y-%m-%d")

dataNew['dateIs'] = factor(sapply(dataNew$date, function(x){ if (weekdays(x) == "domingo" | weekdays(x) == "sabado") { "weekend" } else { "weekday"} }))

#calculate the average of steps w.r.t. the time interval and dateIs
avgStepDateIs = aggregate(steps~interval + dateIs, mean, data=dataNew)

```

Then, we make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). We do this using lattice to look like the one given as example.

``` {r step4 part2}

library(lattice)

xyplot( steps ~ interval | dateIs, data = avgStepDateIs, type="l", layout=c(1,2), xlab="Interval", ylab="Number of steps")

```